---
title: Compare data.table pipes and magrittr pipes
author: GL
date: '2017-07-25'
slug: Compare-data.table-pipes-and-magrittr-pipes
categories:
  - data.table
tags:
  - data wrangling
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

We have two ways to chain `data.table` operations, using `data.table` pipes or using `magrittr` pipes. For a data.table ``dt``, the `data.table` pipes take the form of `dt[][][]...` and `magrittr` pipes `dt %>% .[] %>% .[] %>% ...`. 

Let's first compare the readability of the two pipes in the following examples. Hadley Wickham criticized the readability of ``data.table`` pipes in this [stackoverflow post](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly). The `data.table` pipes, however, are not that hard to follow for those who are familiar with `data.table`. To my eyes, the `magrittr` pipes improve the readability but the `data.table` pipes are still acceptable. It is more of a personal choice.
```{r}
library(data.table)
library(magrittr)

# the data.table pipes
data.table(iris)[
    # add a new column "is_setosa", 1 if yes and 0 if no, use two lines of codes 
    # to clearly show the values
    Species == "setosa", is_setosa := 1
][
    Species != "setosa", is_setosa := 0
][
    # changes the Petal.Length of Species "versicolor", other species not affected
    Species == "versicolor", Petal.Length := 999
][
    # calculate sepal area when length > 5. Area is NA if length <= 5
    Sepal.Length > 5, Sepal.Area := Sepal.Length * Sepal.Width
][
    # select columns
    , .(Species, is_setosa, Petal.Length, Sepal.Area)
][
    # average of each column grouped by species
    , lapply(.SD, mean, na.rm = TRUE), by = Species
]


# the magrittr pipes
data.table(iris) %>%
    # add a new column "is_setosa", 1 if yes and 0 if no, use two lines of codes 
    # to clearly show the values
    .[Species == "setosa", is_setosa := 1] %>%                         
    .[Species != "setosa", is_setosa := 0] %>%
    
    # changes the Petal.Length of Species "versicolor", other species not affected
    .[Species == "versicolor", Petal.Length := 999] %>%   
    
    # calculate sepal area when length > 5. Area is NA if length <= 5
    .[Sepal.Length > 5, Sepal.Area := Sepal.Length * Sepal.Width] %>%  
    
    # select columns
    .[, .(Species, is_setosa, Petal.Length, Sepal.Area)] %>%    
    
    # average of each column grouped by species
    .[, lapply(.SD, mean, na.rm = TRUE), by = Species]                  
```

Will the use of pipes ```%>%``` slow down the computing? In the following code, we add three new columns to a made-up data table with ``data.table`` pipes and `magrittr` pipes. They almost have the same speed.

```{r}
set.seed(123)
dt <- data.table(a = sample(letters, 1e5, replace = TRUE),
                 b = abs(rnorm(1e5)))

datatable_pipe <- function(){
    dt[, x := sqrt(b)][
        , y := b^2
    ][
        , z := paste0(a , b)
    ]
}

magrittr_pipe <- function(){
    dt[, x := sqrt(b)] %>%
        .[, y := b^2] %>%
        .[, z := paste0(a , b)]
}

rbenchmark::benchmark(datatable_pipe(), magrittr_pipe(), replications=20)

```
