---
title: Plot pie charts of racial composition in largest metro areas on a map in R
author: GL Li
date: '2018-01-17'
slug: create-pie-plots-on-a-map-in-r
categories:
  - ggplot
tags:
  - census
---
```{r echo=FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE)
```


Pie chart has been criticized for being a poor visualization and is not recommended in R community. The popular `ggplot2` package discourages the use of pie charts and there is no dedicated `geom_pie` for it.

Although the criticism is mostly valid, there is a case that pie chart can be useful: pie charts on maps. Pie charting on map is a compact way to show composition by locations. 

A recent `R` package, [`scatterpie`](https://github.com/GuangchuangYu/scatterpie) by Guangchuang Yu, specializes in making pie charts at multiple locations. This package is an extension of `ggplot2` so it will be easy for `ggplot2` users.

We will use `scatterpie` package to plot the racial composition of largest metropolitan areas in the United state. The demographics data is extracted from 2016 American Community Survey 1-year estimate with [`totalcensus`](https://github.com/GL-Li/totalcensus) package.

Let's first load the required packages.
```{r}
library(scatterpie)
library(totalcensus)
library(dplyr)
```

Now we extract total population and population of selected races in metro areas from 2016 ACS 1-year estimate. We will only keep metro areas with population over one million. 
```{r}
# read all metro data
metro <- read_acs1year(
    year = 2016,
    states = "US",
    table_contents = c(
        "white = C02003_003", 
        "black = C02003_004", 
        "asian = C02003_006"
    ),
    geo_headers = "CBSA",
    summary_level = "310"
) %>%
    mutate(others = population - white - black - asian)

# keep only largest metro and make some cleaning
largest <- metro %>%
    filter(population > 1e6) %>%
    # Los Angeles metro changed CBSA from 31100 to 31080, hand correct the coordinate
    # mutate(lon = ifelse(GEOID == "31000US31080", -118.18194, lon),
    #        lat = ifelse(GEOID == "31000US31080", 34.10939, lat)) %>%
    select(NAME, lon, lat, total = population, white, black, asian, others) %>%
    arrange(-total)

head(largest, 3)
```

With the race data, we are ready to make pie plot on a map with package `scatterpie`.
```{r}
us <- map_data('state')
ggplot(us, aes(long, lat)) +
    geom_map(map=us, aes(map_id=region), fill="grey97", color="grey") +
    geom_scatterpie(data = largest, 
                    aes(lon, lat, r = sqrt(total)/2000),
                    cols = c("white", "black", "asian", "others"), 
                    alpha = 0.5) +
    scale_fill_manual(
        breaks = c("white", "black", "asian", "others"),
        labels = c("white alone", "black alone", "asian alone", "others"),
        values = c("asian" = "orange",
                   "white" = "white",
                   "black" = "black",
                   "others" = "cyan")
    ) +
    labs(title = "Races in metro areas over one million population",
         subtitle = "others includes mixed races and American indians, Alaska natives, native Hawaiians and pacific islanders",
         caption = "Source: 2016 ACS 1-year estimate",
         fill = NULL) +
    coord_fixed() +
    theme_bw() +
    theme(legend.position = c(0.96, 0.02),
          legend.justification = c(1, 0),
          panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank())

```

