---
title: My uniform way of using ggplot2
author: GL Li
date: '2017-08-03'
slug: a-uniform-way-to-use-ggplot2
categories:
  - ggplot
tags:
  - plot
---

```{r echo=FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE)
```

Just finished reading Hadley Wickham's ``ggplot2`` book, ([second eition](http://www.springer.com/us/book/9783319242750)). Before that I have been using ``ggplot2`` for a couple of years, mainly learned by reading documentation and searching for help online. 

The `ggplot2` package is a powerful and comprehensive tool for generating static plots, but I also feel it is a little bit too flexible; the same plot can be made with many different ways. This fexibility provides obvious convenience but also introduces a lot of confusion and extra burden of memorization. Being flexible is not a good feature for a lazy user like me. I just can not memorize all these different methods. What I want is an appoach that gets the job done, is easy to apply to many tasks, and does not have much to memorize. I need a uniform way to use `ggplot2`.

Here I discuss how I like to make `ggplot2` plots. It is my personal preference. Every one can find their best way. I will give the reason why I am doing so. 

## What to be included in ``ggplot()``
Every `ggplot2` plotting starts with the function ``ggplot()``. Typically `data` and `aes()` are included as arguments. The `aes()` can include `x`, `y` and more aesthetic mapping such as `color`, `shape`, `fill`, ..., as in the code below,

```{r eval=FALSE}
ggplot(data = mpg, aes(x = displ, y = cty, color = drv, shape = year, fill = cyl))
# or to simplify
ggplot((mpg, aes(displ, cty, color = drv, shape = year, fill = cyl)))
```

My preference for this part is to include as fewer arguments as possible, even without any argument. All these arguments are inherited by all plotting layers by default. If you want them to be used in all layers, then that is good; you save a few keystrokes. But if it is not what you want, it is where troubles hide. For example, if your want to a linear fit to all data points but has `color` in aesthetic mapping, the fit is applied to each color. You'd better move the `color` out of `ggplot()`. 

```{r fig.height=3, fig.width=5}
library(ggplot2)
ggplot(mpg, aes(cty, hwy, color = drv)) +
    geom_point() +
    stat_smooth(method = "lm")
```


## Which to choose, `geom_xxx()` or `stat_xxx()`
These two functions caused a lot of headache to me. Both `geom_xxx()` and `stat_xxx()` plot layers. Many `geom`s have equivalent `stats`, such as `geom_smooth()` and `stat_smooth()`, `geom_bar()` and `stat_count`, and `geom_boxplot()` and `stat_boxplot()`. Even `geom_point()` has an equivalent `stat_identity()`. So which one to use? 

I like to use `geom_xxx()` for layers that directly plot the original data and `stat_xxx()` for layers that plot statistical transformation of the original data. The former includes `geom_point()`, `geom_line()`, and `geom_col()` etc. The latter includes `stat_smooth()`, `stat_count`, and `stat_boxplot()` etc. This a natural choice to me as I pay more attention to data; whenever statistical transformed data are used, I want to use it explicitly with function starting with `stat`. Using `stat_xxx()` has additional benefit: we can choose the right `geom` as desired without losing sense of what data are being displayed. In the following code, the counts of `drv` are displayed as bar, point and line. No matter what `geom` you choose, the `stat_count` always explicitly tells you that you are plotting the count. 

```{r fig.height=3, fig.width=4}
ggplot(mpg, aes(drv)) +
    stat_count(geom = "bar", fill = "grey70") +
    stat_count(geom = "point", size = 6, color = "blue") +
    stat_count(geom = "line", aes(group = 1), color = "red")
```

A few words about bar plot. Above count of one categorical variable are often plotted with `geom_bar()`. The awkward thing is that `geom_bar(stat = "identity")` also make bar plot when two variables `aes(x, y)` are provided. To solve this issue, `ggplot2` version 2.2.0 introduced `geom_col(aes(x, y))` to take care bar plot from two variables. With `stat_count()` and `geom_col()`, `geom_bar()` are ready to retire.


## How to organize axis and legend
Almost all components regarding axis and legend can be specified in `scale_xxx_xxx()` funtions. You can get the job done by only using them. The code may be too busy though. Several help functions do part of `scale_xxx_xxx()`'s job and make the code succint, such as `xlab()` and `ylim()`. But too many help functions also make code less structured.

I like two help functions: `labs()` sets titles of all axis and lengends and `guides()` arranges multiple legends. Together with `scale_xxx_xxx()`, the three take care axis and legend in a logical way:

- `scale_xxx_xxx()` determines what to show. They are used to set limits, breaks and labels of each aesthetics, which are the real data to show.
- `guides()` determines how to display. It is a best practice to arrange each legend by considering all others the same time. 
- `labs()` determines how to call them. It names axis and legend in one place by treating all axis and legend as a whole, which helps to name them in the same style. 

```{r}
ggplot(mpg, aes(hwy, cty, color = factor(cyl), shape = drv)) +
    geom_point(aes(size = displ)) +
    # what to display
    scale_x_continuous(limits = c(0, 50),
                       breaks = c(0, 20, 40),
                       labels = c("zero", "twenty", "fourty"), 
                       minor_breaks = c(5, 10, 15)) +
    scale_color_discrete(breaks = c(4, 5, 6, 8), 
                         labels = c("four", "five", "six", "eight")) + # asign colors by default
                                    
    scale_shape_manual(limits = c("f", "r"),      # manually select shape
                       values = c(f = 0, r = 2)) +
    # how to display
    guides(
        size = "none",  # hide size
        color = guide_legend(direction = "horizontal",
                             title.position = "top",
                             nrow = 2,
                             byrow = TRUE,  # arrange row by row, default is by column
                             order = 1),    # first legend to show
        shape = guide_legend(direction = "vertical",
                             reverse = TRUE)  # reverse order of legend keys
    ) +
    # what's your name
    labs(title = "Use scale_xxx(), guides(), and labs() for legend and axis",
         subtitle = "scale_xxx for what to display, guides for how to display, labs for how to name",
         x = "highway mileage",
         y = "city mileage", 
         color = "cylinder", 
         shape = "drive train", 
         size = "haha you cannot see me") 

```

## Let's do a quick annotation with `annotate()`
Annotation with ``annotate()`` is handy when you simply want to draw something using vectors. It can draw `text`, `segment`, `rect`, and pretty much everything `geom_xxx()` offers. Do not get confused with `geom_text()` and `geom_segment()`, and `geom_rect`, ..., which use dataframes. Reserve `annotate()` for drawing simple things manually.

```{r}
ggplot(mpg, aes(displ, hwy, color = drv)) +
    geom_point() +
    annotate("text", x = 6, y = 35, label = "These big guys\nare gas efficient.") +
    annotate("segment", x = c(5.8, 6, 6.2), y = 32.5, 
             xend = c(5.7, 6.15, 6.9), yend = c(27, 27, 25.1),
             arrow = arrow(angle = 20, length = unit(2, "mm"), type = "closed"))
```

## Finally a comprehensive example
In addition to the confusing parts discussed above, a normal `ggplot2` plot may contains `position`, `facet`, and of course `theme()`. I will conclude this post with an example that covers all major components of a `ggplot2` plot. This is the uniform way I am using `ggplot2`.

```{r}
ggplot(mpg, aes(factor(cyl), cty)) +
    # leave color and size in geom_point() so that stat_boxplot works on all data
    geom_point(aes(color = drv, size = displ), position = "jitter") +
    stat_boxplot(fill = NA) + 
    facet_wrap(~year) +

    # === take care axis and legend ===
    
    # determine what values and labels to display
    scale_x_discrete(breaks = c(4, 5, 6, 8),
                     labels = c("Four", "Five", "Six", "Eight")) +
    scale_y_continuous(limits = c(4, 36),
                       breaks = 1:4 * 8) +
    scale_color_manual(breaks = c("4", "f", "r"),
                       values = c("4" = "red", "f" = "blue", "r" = "cyan"),
                       labels = c("four-wheel", "front-wheel", "rear-wheel")) +
    scale_size_area(limits = c(2, 7),
                          max_size = 4) +
    
    # make arrangement for each legend
    guides(
        size = guide_legend(
            title.position = "top",
            nrow = 2,
            byrow = TRUE,
            override.aes = list(shape = 1),
            reverse = TRUE
        ),
        color = guide_legend(
            order = 1,
            nrow = 3,
            override.aes = list(size = 3)
        )
    ) +
    
    # using labs() for all of the titles and labels in one place
    labs(title = "This plot displays all major ggplot components",
         subtitle = "including data, aes mapping, geom, stat, position, facet, scale, guides, annotaion, and theme",
         caption = "Source: what so ever",
         x = "Cylinders",
         y = "City Mileage (miles/gallon)",
         color = NULL,
         size = "displacement") +
    
    # === provide extra information ===
    
    # annotate extra geoms mannually. Play with geom_text() if want to label in
    # a specific facet panel
    annotate("text", x = 0.5, y = 6, label = "boxplot is applied to all data", 
             hjust = 0, vjust = 1) +
    
    # === use theme() to make it beautiful ===
    
    theme(plot.background = element_rect(fill = "#F5E6E3"),
          plot.title = element_text(family = "monospace"),
          plot.subtitle = element_text(face = "italic"),
          panel.background = element_rect(fill = "lightblue", color = "red"),
          panel.grid.major.y = element_line(color = "grey95", size = 0.2),
          panel.grid.minor.y = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(family = "monospace"),
          legend.position = "top",
          legend.key = element_blank(),
          legend.margin = margin(0, 0, 0, 0),
          legend.background = element_blank(),
          strip.background = element_blank(),
          strip.text = element_text(size = 12))
```

