---
title: "ggplot2: place text at right location"
author: GL Li
date: '2017-08-18'
slug: place-text-at-right-location
categories:
  - ggplot
tags:
  - plot
---

```{r echo=FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE)
```

A common task in plotting is adding texts as labels or annotations to specific locations. `ggplot()` has functions `geom_text()`, `geom_label()` and `annotate()` for this purpose. In this post we discuss how `ggplot2` controls positioning of text.

First we need to specify `(x, y)` coordinate in the plot where the text is placed. By default, the center of the text is at `(x, y)`, which is sometimes not what we want, as shown in the example below. 

```{r fig.width=5, fig.height=3}
library(ggplot2)
df <- data.frame(x = 1:3, 
                 y = 1:3, 
                 z = c("aa\nnew line", "bb\nnew line 1\nand new line 2", "cc"))
g <- ggplot(df, aes(x = x, y = y)) +
    geom_col() +
    geom_point(color = "red") +
    coord_cartesian(xlim = c(0.3, 3.5), ylim = c(0.5, 3.5))
g + geom_text(aes(label = z)) +
    labs(subtitle = "(x, y) set the position of text but often is not we want")
```


As the text has its own shape and size, we need to determine which part of the text to be superimposed at `(x, y)`. This positioning is controlled by parameters `hjust` and `vjust`. Imaging the text is surrounded by a rectangle and a local coordinate originates at the bottom left corner of the rectangle, `(hjust, vjust)` then is a local position in the local coordinate. By default, `hjust = 0.5` and `vjust = 0.5`, that is, the center of the rectangle is at `(x, y)`.  We can change them to match to `(x, y)` by borders or corners, as shown in the examples below. The positioning of text is still not ideal as it is too close to `(x, y)`.

```{r fig.width=10, fig.height=3}
library(grid)
library(gridExtra)

g1 <- g + geom_text(aes(label = z), hjust = 0) +
    labs(subtitle = "hjust = 0 place left side of the text to (x, y)")
g2 <- g + geom_text(aes(label = z), vjust = 1) +
    labs(subtitle = "vjust = 1 place top of the text to (x, y)")
g3 <- g + geom_text(aes(label = z), hjust = 1, vjust = 0) +
    labs(subtitle = "hjust = 1 and vjust = 0 place tottom\nright cornor of the text to (x, y)")

grid.arrange(g1, g2, g3, nrow = 1)
```

The parameters `nudge_x` and `nudge_y` shift whole text along `x` and `y` axis. With them we can add a space between text and `(x, y)`. 

```{r fig.width=10, fig.height=3}
g1 <- g + geom_text(aes(label = z), hjust = 0, nudge_x = 0.1) +
    labs(subtitle = "nudge_x = 0.1 moves text to the right by 0.1")
g2 <- g + geom_text(aes(label = z), vjust = 1, nudge_y = -0.1) +
    labs(subtitle = "nudge_y = -0.1 moves text down by 0.1")
g3 <- g + geom_text(aes(label = z), hjust = 1, vjust = 0, nudge_x = -0.1, nudge_y = 0.1) +
    labs(subtitle = "nudge_x = -0.1 nudge_y = 0.1 to northwest")

grid.arrange(g1, g2, g3, nrow = 1)
```

For multi-line text, we also want to adjust the line space. The line space is set by parameter `lineheight`. 

```{r fig.width=10, fig.height=3}
g1 <- g + geom_text(aes(label = z), hjust = 1, vjust = 0, nudge_x = -0.1, lineheight = 0) +
    labs(subtitle = "lineheight = 0")
g2 <- g + geom_text(aes(label = z), hjust = 1, vjust = 0, nudge_x = -0.1, lineheight = 0.9) +
    labs(subtitle = "lineheight = 0.9")
g3 <- g + geom_text(aes(label = z), hjust = 1, vjust = 0, nudge_x = -0.1, lineheight = 2) +
    labs(subtitle = "lineheight = 2")

grid.arrange(g1, g2, g3, nrow = 1)
```

As a final note, some people use `hjust` and `vjust` that are `< 0` or `> 1` to shift text from `(x, y)`. This works well to move up and down a single line text but not good to move left and right or multi-line text, as `hjust` and `vjust` are relative to dimension of the rectangle around the text. Some bad examples are shown below. So it is better to use `nudge_x` and `nudge_y` to shift whole text.

```{r fig.width=8, fig.height=3}
g1 <- g + geom_text(aes(label = z), vjust = 1.3) +
    labs(subtitle = "vjust = 1.3, more line more space")
g2 <- g + geom_text(aes(label = z), hjust = -0.2) +
    labs(subtitle = "hjust = -0.1, longer lines more space")

grid.arrange(g1, g2, nrow = 1)
```

